Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-01-17 acm/ieee_paper.tex
Incorporated Evan's changes. Blue is stuff I change (most of the stuff I cahnged). Red are questions.

In this paper, we consider the case of local stencils with a fixed number
of nodes at each point and encode the sparse matrix in ELLPACK
format. We increase the number of operations relative to memory
bandwidth by \ge{interleaving} the calculation of  four derivatives 
of four different functions,
or 16 different derivatives. We demonstrate a novel implementation on
the Intel MIC architecture, taking into account its advanced swizzling and
channel interchange features. We present benchmarks on a real data
set that show an almost eightfold increase in speed compared to
efficient implementations of a single derivative, reaching a
performance of almost 140~Gflop/s. We explain the results through
consideration of operation count versus memory bandwidth.
\end{abstract}

@@

Another solution, pursued in this paper, \ge{is to compute multiple SpMVs using matrices with identical spasity patterns, but with different matrix elements.}  Obviously not all the applications have such a property. However, important classes of applications such as graph recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of derivatives for solving systems of PDEs using Radial Basis Function-generated Finite Differences (RBF-FD)~\cite{FLBWSC12} \ge{fall into the category of applications that require multiple SpMVs simultaneously.} RBF-FD is a meshless method, which easily handles irregular geometries and local refinement with algorithmic complexity independent of dimensionality, 
and which can produce high-order derivative approximations. These methods are rapidly gaining 
ground in science and engineering modeling communities \cite{Bayona13,CDNT,FoL11,FLBWSC12,SPLM}). 
As a result, it is of interest to develop an efficient implementation on novel computer plat
forms for the calculation of derivatives within the context of RBF-FD, which thus far are responsible 
for the bulk of computer resources. \ge{Typical equations solved by RBF-FD depend on multiple derivatives
of multiple functions. Rather than compute each derivative individually, this paper investigates 
how to compute all derivatives simultaneously in an interleaved fashion. Specifically, 
we calculate four different derivatives (corresponding to four different sparse matrices with identical 
sparsity pattern) of four different functions
%is an interThis paper investigates the possibility  to increase the ratio  case of approximating 
%four different derivatives, using RBF-FD, for four different functions 
(a common scenario in 3D fluid dynamics modeling) for a total of 16 derivatives. }
%Since the RBF-FD differentiation matrices approximating the four different derivative operators have an identical sparsity pattern, this leads to the simultaneous execution of 16 SpMVs at a time. Using multiple vectors at a time has been investigated before to densify the computations~\cite{Im01} but, to the best of the authors’ knowledge, this is the first time that the densification comes from adding both vectors and matrices.

@@

Section~\ref{sec:model} presents an
estimation of the instruction intensity of various forms of the
computations. We show that a sixfold improvement can be
expected when computing the sixteen multiplications simultaneously to
reach a total of about 210~Gflop/s. This performance represents
approximately 10\% of the available flop/s of a Xeon Phi
coprocessor. \ge{Therefore, it is necessary to have an implementation that
performs the computation in as few instructions as
possible.} \NOTE{I DO NOT FOLLOW the Therefore in the previous sentence.}
\ge{We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instructions to efficiently bring the data from memory into the vector
registers to be processed. Section~\ref{sec:expe} presents results on 
memory access and computational speed and their relation to choices made in the 
SpMV kernels.}
%experimental results  discussing the relationship between the structure of $A$ and 
%the ; specifcally on the amount of bandwidth that can be achieved
%depending on how the SpMV kernel is written. 
It also provides the actual performance of the various kernels on multiple classes of matrices, 
some generated for purpose of analysis, and some extracted from an application of RBF-FD. 
A performance of 135~Gflop/s is achieved using RBF-FD differentiation matrices. Concluding remarks and perspectives are provided in Section~\ref{sec:ccl}.\looseness=-1

@@

Each core in the architecture has a 32~kB L1 data cache, a 32~kB L1
instruction cache, and a 512~kB L2 cache. The architecture of a core is
based on the traditional Pentium architecture, which has been extended to
64-bit. A core can hold four hardware contexts at any given time (240 threads in total). At each
clock cycle, instructions from a single thread are executed. Due to
some hardware constraints, two hardware contexts must be used to reach
the peak instruction throughput in that architecture. Similar to the
Pentium architecture, a core has two different concurrent instruction
pipelines that allow the execution of two instructions per
cycle. However, only one vector or floating point instruction can be
executed per cycle.

@@

\begin{figure*}
  \centering
\subfigure[Source code]{\includegraphics[clip=true,trim=4.5cm 11cm 4cm 4.2cm,width=.6\linewidth]{figures/code.pdf}}%
\subfigure[Content of the vector registers]{\hspace{-0.07\linewidth}\includegraphics[width=.5\linewidth]{figures/code-annote.pdf}}

@@

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the No Read No Global
Order hint to optimize write traffic on the memory bus.

@@

The {\tt read, cpp} implementation reaches a performance of 168~GB/s
 which is better than the {\tt read}
implementation which only achieves 145~GB/s. This
is explained by the compiler adding loop unrolling and software
prefetching in the assembly code in the {\tt read, cpp} case, which is
not added in the {\tt read} case. On the other hand, the {\tt write}
case reaches a performance of 158~GB/s while the
{\tt write, cpp} implementation never exceeds 58~GB/s. 
This difference is due to the use of the {\tt \_mm512\_store\_nrngo\_ps} instruction, 
which bypasses the Read For Ownership protocol and allows the writes
to be performed in any order. The compiler can not use this
instruction on its own because this would break the memory convention
useful for some parallel algorithms. The {\tt read/write} benchmark
performs at 140~GB/s while the {\tt read/write, cpp} benchmark performs
at about 90~GB/s.

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses is removed from consideration. The cost associated with 
{\tt A} remains. The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the Random
matrices (Figure~\ref{fig:compact}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache-based architecture where the cache
reutilisation is the lowest from one row to the next.
The other type of matrix is used in derivative stencils in 3D
RBFFD calculations (Figure~\ref{fig:rbf3dnorcm} shows a $32$-points
stencil of a 3D $8^3$ grid).

@@

  \caption{Performance of $y=Ax$ for a RBF derivative stencil of a
    $96^3$ 3D grid. Classical SpMV ($n_m=n_v=1$ given in blue) has low
    performance. The manually vectorized code (red) is 3.5 times
    faster than the compiler generated one (green). RCM ordering
    (squares) only improves the performance for our manually
    vectorized code.}
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
\end{figure}

@@

To put things in perspective, we conducted some experiments on a
classical CPU architecture with two Intel Core i7 processors, each has eight
cores that support AVX instructions. Both the compiler-generated and
the manually-written vectorization using AVX instructions reach a
(read) bandwidth of 70~GB/s. This indicates that the performance of
the classical SpMV can not overcome 17~Gflop/s. A generic
implementation of the multiplication of four vectors with four matrices on
the {\tt RBF} matrix reaches a performance of 27~Gflop/s; while a
manual implementation using AVX instructions reaches a performance of
42~Gflop/s. This confirms that our method is sound and generalizes to
architectures other than the Intel MIC architecture.

@@

Our implementation makes use of the IMCI MIC instruction set and
includes a number of swizzling and channel swapping operations, for an
extremely efficient tensor product implementation. On a $96^3$ 3D
grid, it reaches a performance of 140~Gflop/s which is 2.7 times
faster than the best possible performance achievable for a single
vector and a single derivative and more than 7 times faster than a
practical implementation.

>>>>

In this paper, we consider the case of local stencils with a fixed number
of nodes at each point and encode the sparse matrix in ELLPACK
format. We increase the number of operations relative to memory
bandwidth by \ge{interleaving} the calculation of  four derivatives 
of four different functions,
or 16 different derivatives. We demonstrate a novel implementation on
the Intel MIC architecture, taking into account its advanced swizzling and
channel interchange features. We present benchmarks on a real data
set that show an almost \ge{sevenfold} \note{eightfold; there were inconsistencies in the ppaer.} increase in speed compared to
efficient implementations of a single derivative, reaching a
performance of almost 140~Gflop/s. We explain the results through
consideration of operation count versus memory bandwidth.
\end{abstract}

@@

Another solution, pursued in this paper, \ge{is to compute multiple SpMVs using matrices with identical spasity patterns, but with different matrix elements.}  Obviously not all the applications have such a property. However, important classes of applications such as graph recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of derivatives for solving systems of PDEs using Radial Basis Function-generated Finite Differences (RBF-FD)~\cite{FLBWSC12} \ge{fall into the category of applications that require multiple SpMVs simultaneously.} RBF-FD is a meshless method, which easily handles irregular geometries and local refinement with algorithmic complexity independent of dimensionality, 
and which can produce high-order derivative approximations. These methods are rapidly gaining 
ground in science and engineering modeling communities \cite{Bayona13,CDNT,FoL11,FLBWSC12,SPLM}). 
As a result, it is of interest to develop an efficient implementation on novel computer platforms for the calculation of derivatives within the context of RBF-FD; derivative calculations that account for the bulk of computer resources when running a numerical simulation implemented with RBF-FD. 
\ge{Typical equations solved by RBF-FD depend on multiple derivatives
of multiple functions. Rather than compute each derivative individually, this paper investigates 
how to compute all derivatives simultaneously in an interleaved fashion. Specifically, 
we calculate four different derivatives (corresponding to four different sparse matrices with identical 
sparsity pattern) of four different functions
%is an interThis paper investigates the possibility  to increase the ratio  case of approximating 
%four different derivatives, using RBF-FD, for four different functions 
(a common scenario in 3D fluid dynamics modeling) for a total of 16 derivatives. }
%Since the RBF-FD differentiation matrices approximating the four different derivative operators have an identical sparsity pattern, this leads to the simultaneous execution of 16 SpMVs at a time. Using multiple vectors at a time has been investigated before to densify the computations~\cite{Im01} but, to the best of the authors’ knowledge, this is the first time that the densification comes from adding both vectors and matrices.

@@

Section~\ref{sec:model} presents an
estimation of the instruction intensity of various forms of the
computations. We show that a \ge{sevenfold} \NOTE{sixfold was inconsistent with eighfold in the abstract.} improvement can be
expected when computing the sixteen multiplications simultaneously to
reach a total of about 210~Gflop/s. This performance represents
approximately 10\% of the available flop/s of a Xeon Phi
coprocessor. \ge{Therefore, it is necessary to have an implementation that
performs the computation in as few instructions as
possible.} \NOTE{I DO NOT FOLLOW the Therefore in the previous sentence. Evan has the same concern.}
\ge{We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instructions to efficiently bring the data from memory into the vector
registers to be processed. Section~\ref{sec:expe} presents results on 
memory access and computational speed and their relation to choices made in the 
SpMV kernels.}
%experimental results  discussing the relationship between the structure of $A$ and 
%the ; specifcally on the amount of bandwidth that can be achieved
%depending on how the SpMV kernel is written. 
It also provides the actual performance of the various kernels on multiple classes of matrices, 
some generated for purpose of analysis, and some extracted from an application of RBF-FD. 
A performance of 135~Gflop/s is achieved using RBF-FD differentiation matrices. Concluding remarks and perspectives are provided in Section~\ref{sec:ccl}.\looseness=-1

@@

Each core in the architecture has a 32~kB L1 data cache, a 32~kB L1
instruction cache, and a 512~kB L2 cache. The architecture of a core is
based on the traditional Pentium architecture, which has been extended to
64-bit with hyperthreading. A core can hold four hardware contexts at any given time (240 threads in total). At each
clock cycle, instructions from a single thread are executed. Due to
some hardware constraints, two hardware contexts must be used to reach
the peak instruction throughput in that architecture. Similar to the
Pentium architecture, a core has two different concurrent instruction
pipelines that allow the execution of two instructions per
cycle. However, only one vector or floating point instruction can be
executed per cycle.

@@

\begin{figure*}
  \centering
\subfigure[Source code]{\includegraphics[clip=true,trim=4.5cm 11cm 4cm 4.2cm,width=.6\linewidth]{figures/code.pdf}}%
\subfigure[Content of the vector registers \ge{in the source code in (a)}.]{\hspace{-0.07\linewidth}\includegraphics[width=.5\linewidth]{figures/code-annote.pdf}}

@@

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the "No Read No Global
Order" hint to optimize write traffic on the memory bus.

@@

The {\tt read, cpp} implementation reaches a performance of 168~GB/s
 which is better than the {\tt read}
implementation which only achieves 145~GB/s. This
is explained by the compiler adding loop unrolling and software
prefetching in the assembly code in the {\tt read, cpp} case, which is
not added in the {\tt read} case. On the other hand, the {\tt write}
case reaches a performance of 158~GB/s while the
{\tt write, cpp} implementation never exceeds 58~GB/s. 
This difference is due to the use of the {\tt \_mm512\_store\_nrngo\_ps} instruction, 
which bypasses the \ge{"Read For Ownership"} protocol and allows the writes
to be performed in any order. The compiler can not use this
instruction on its own because this would break the memory convention
useful for some parallel algorithms. The {\tt read/write} benchmark
performs at 140~GB/s while the {\tt read/write, cpp} benchmark performs
at about 90~GB/s.

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
\ttt{Supercompact} matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses is removed from consideration. The cost associated with 
{\tt A} remains. The \ttt{Compact} matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the \ttt{Random}
matrices (Figure~\ref{fig:compact}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache-based architecture where the cache
reutilisation is the lowest from one row to the next.
The other type of matrix is used in derivative stencils in 3D
RBFFD calculations (Figure~\ref{fig:rbf3dnorcm} shows a $32$-point
stencil of a 3D $8^3$ grid).

@@

  \caption{Performance of $y=Ax$ for a RBF derivative stencil of a
    $96^3$ 3D grid. Classical "one-matrix one-vector" SpMV ($n_m=n_v=1$ given in blue) has low
    performance. The manually vectorized code (red) is 3.5 times
    faster than the compiler generated one (green). RCM ordering
    (squares) only improves the performance for our manually
    vectorized code.}
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
\end{figure}

@@

To put things in perspective, we conducted similar experiments on a
classical CPU architecture with two Intel Core i7 processors, each has eight
cores that support AVX instructions. Both the compiler-generated and
the manually-written vectorization using AVX instructions reach a
(read) bandwidth of 70~GB/s. This indicates that the performance of
the classical SpMV can not overcome 17~Gflop/s. A generic
implementation of the multiplication of four vectors with four matrices on
the {\tt RBF} matrix reaches a performance of 27~Gflop/s; while a
manual implementation using AVX instructions reaches a performance of
42~Gflop/s. This confirms that our method is sound and generalizes to
architectures other than the Intel MIC architecture.

@@

Our implementation makes use of the IMCI MIC instruction set and
includes a number of swizzling and channel swapping operations, for an
extremely efficient tensor product implementation. \NOTE{From Evan: this is the first time $96^3$ is mentionned. Might wish to mention it earlier.} On a $96^3$ 3D
grid, it reaches a performance of 140~Gflop/s which is 2.7 times
faster than the best possible performance achievable for a single
vector and a single derivative and more than 7 times faster than a
practical implementation.
